Combined effects of protein expression variance and correlation on multicomponent systems

### 2020-03-11 Manuscript/Manuscript.Rmd
.

  word_document: null
documentclass: article
capsize: normalsize
geometry: margin=1in
fontsize: 11pt
spacing: doublespacing
bibliography: 
  - references.bib
  - knitcitations.bib
csl: cell.csl
abstract: Protein expression variation leads to phenotypic variance between cells. This has been demonstrated in cell signaling and differentiation decisions. Additionally coordinated expression of proteins between cells can tune signaling pathways either towards a more binary or analog modality. Though there is some evidence in bulk cell measurements and in bacteria that certain heteromeric subunits or metabolic pathways may be expressed in a coordinated fashion (i.e. operons in bacteria), there has not been a direct measurement of coordinated expression of proteins (independent of TFs) in metazoans (vertebrates?). Here, we measure cell-to-cell variability of relative protein abundance using quantitative proteomics of individual *Xenopus laevis* eggs and show that proteins involved in metabolic pathways or members of heteromeric complexes tend to have high correlations with other members of those pathways/complexes. Our previous work highlighted the fact that correlated expression increases the total variation of a pathway, so one would reason that certain pathways or complexes would need to compensate of this extra source of by reducing the variation of expression of these pathways or complexes. To test this we computed total variance score that took into account both the coefficient of variance and correlations between proteins in a pathway and found that the lower 10% of GO terms were highly enriched for metabolic pathways. When we looked at the relationship between CV and R between these GO terms we found a negative relationship between them, demonstrating that increased correlation needs to come at the expense of decreased variance. Simple molecular models of heteromeric complexes and metabolic pathways demonstrate that this tradeoff can result in higher efficiencies in both function and reduced energy waste. Together, our study argues for a control principle whereby coordinated expression of proteins in a pathway can require lower variance in order to reduce the overall pathway variance, enabling accurate control of active complexes or metabolic pathway activity.
---

@@

library(knitr)
library(tidyverse)
library(RColorBrewer)
#library(pheatmap)
library(ComplexHeatmap)
library(GGally)
library(VIM)
library(org.Xl.eg.db)
library(GO.db)
library(ggrepel)
library(plotly)
library(cowplot)
library(patchwork)
source("../functions.R")
library(knitcitations)
library(ggplotify)
library(latex2exp)
library(KEGG.db)

@@

```{r group_prots, include=FALSE}
# GO Terms
GO <- protImpute %>% dplyr::select(Uniprot) %>% filter(Uniprot %in% keys) %>% 
  mutate(GOterms = mapIds(org.Xl.eg.db, 
                     keys = Uniprot, 
                     keytype = 'UNIPROT', 
                     column = 'GOALL', 
                     multiVals = 'list'))

@@

allKEGG <- unnest(kegg, cols = KEGG) %>% nest(Uniprot = Uniprot) %>%
  rowwise() %>% mutate(n = length(unlist(Uniprot))) %>% ungroup() %>%
  mutate(name = kegg_names$kegg_name[match(KEGG,kegg_names$kegg_id)],
         type = "KEGG",
         ontology = NA) %>% rename(accession = KEGG) %>%
  dplyr::select(type, accession, name, ontology, Uniprot, n)

@@

```{r total_var, include=FALSE}
groupCor <- function(x){
  geneList <- unlist(x)

@@

allGroups <- allGroups %>% filter(!is.na(accession)) %>% rowwise() %>%
  mutate(totVar = totVarGroup(Uniprot),
         normVar = totVar / n) %>%
  filter(n > 3) %>% ungroup()
```

@@

```{r figure_1, echo=FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, fig.cap=fig_legend_1}

@@

```{r figure_2,, echo=FALSE, fig.height=8, fig.width=6, message=FALSE, warning=FALSE, fig.cap=fig_legend_2}
library(magick)

p1 <- ggplot(allGroups, aes(x = normVar, fill = interval)) + 
  geom_histogram(bins = 50) + theme_minimal() +
  #scale_fill_viridis_d()
  scale_fill_manual(values = c("#d53e4f","#3288bd","grey")) +
  xlab("Normalized Variation") + ylab("Number")

interesting_groups_low <- c("03010","00030","GO:0046496","00071",
                            "GO:0006096","GO:0015934","GO:0019843",
                            "00051","GO:0042555","04012")
interesting_groups_high <- c("GO:0042995","03008","GO:0007264","GO:0005885","GO:0007166","04150","GO:0035556","GO:0016791","GO:0007165")
interesting_groups <- c(interesting_groups_low, interesting_groups_high)

p2 <- ggplot(filter(allGroups, n > 5), aes(x = medCors, y = medCV, color = interval)) + 
  geom_point(data = filter(allGroups, interval == "mid", n > 4),
             size = 2, alpha = 0.5, stroke = 0) +
  geom_point(data = filter(allGroups, interval != "mid", n > 4),
             size = 3, alpha = 0.75, stroke = 0) +
  geom_label_repel(data = filter(allGroups, accession %in% interesting_groups), 
                  aes(label = name), 
                  colour = 'black', 
                  size = 2,
                  force = 5,
                  box.padding   = 1) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_color_manual(values = c("#d53e4f","#3288bd","grey")) +
  xlab("Mean Correlation") + ylab("Mean Coefficient of Variation")

@@

p5 <- rbind(mutate(allGroups, interval = "mid", name = "all"),
      filter(allGroups, accession %in% interesting_groups)) %>%
  select(-Uniprot,-covs,-cv,-type,-ontology) %>%
  unnest(cols = "cors") %>%
  ggplot(., aes(x = fct_reorder(name,rank(interval)), y = cors, fill = interval)) + 
  geom_boxplot(notch = TRUE, alpha = 0.75) +
  scale_fill_manual(values = c("#d53e4f","#3288bd","grey")) +
  coord_flip() + 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(size = 8),
        axis.title.y=element_blank())

p6 <- p2 + p5 + plot_layout(widths = c(1,0.5))

(p1 + p3 + p4) / p6 + plot_layout(heights = c(1,4)) +
  plot_annotation(tag_levels = "A") & 
  theme(legend.position = "none")

fig_legend_2 <- c("A) Histogram of the normalized variation values across all of the GO terms and KEGG pathways. The bottom and top 10% are colored blue and red respectively. B) Pairwise plot between the mean correlation and mean coefficient of variaiton within each GO term and KEGG pathway with color coding from A. Interestingly, both the top and bottom 10% of normalized variation have groups of proteins with high levels of correlation, despite high correlation adding to total variance. In the lowest 10% of normalized variation groups there appears to be a negative correlation between mean correlation and variation.")
```

```{r sup_fig_1, echo=FALSE, fig.height=2, fig.width=6, message=FALSE, warning=FALSE, fig.cap=sup_fig_legend_1}
allGroups %>% filter(n > 4) %>%
  ggplot(., aes(x = medCors, y = medCV, color = interval)) + 
  geom_point() + 
  scale_color_manual(values = c("#d53e4f","#3288bd","grey")) +
  stat_summary(fun.data=mean_cl_normal) + 
  geom_smooth(method='lm', formula= y~x, color = "black") +
  facet_wrap(~interval, scales = "free", nrow = 1) +
  theme_minimal() +
  theme(legend.position = "none")

```{r figure_3, echo=FALSE, fig.height=8, fig.width=6, message=FALSE, warning=FALSE, fig.cap=fig_legend_3}
library(circlize)

allGroups[grep("large ribo",allGroups$name),]
prots <- allGroups %>% filter(accession == "GO:0015934") %>% pull(Uniprot) %>% unlist() %>% as.vector()
mat <- cor[prots,prots]
rownames(mat) <- mapIds(org.Xl.eg.db, keys = rownames(mat), keytype = 'UNIPROT', column = 'SYMBOL')

col_fun = colorRamp2(seq(-1,1,0.01), rev(colorRampPalette(brewer.pal(11, 'RdBu'))(201)))

p1 <- as.ggplot(
  Heatmap(
    mat,
    column_title = "Large ribosomal subunit",
    column_title_gp = gpar(fontsize = 20, fontface = "bold"),
    col = col_fun,
    heatmap_legend_param = list(title = "R"),
    show_column_dend = TRUE,
    show_row_dend = TRUE,
    show_column_names = TRUE,
    show_row_names = TRUE,
    use_raster = FALSE,
    cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%.1f", mat[i, j]), x, y, gp = gpar(fontsize = 10))}
  )
)

@@

Heatmap(totVarMat,
        col = col_fun,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        border = NA,
        column_title = "R",
        row_title = "CV",
        heatmap_legend_param = list(title = "Norm Var"),
        row_names_gp = gpar(fontsize = 5),
        column_names_gp = gpar(fontsize = 5))

@@

$$
\begin{bmatrix} 
\ \sigma(X_{1})^2 +  \rho(X_1,X_2) \cdot \sigma(X_{1}) \cdot \sigma(X_{2}) \\
\sigma(X_{1})^2 + \rho(X_1,X_3) \cdot \sigma(X_{1}) \cdot \sigma(X_{3}) \\
\sigma(X_{1})^2 + \rho(X_1,X_4) \cdot \sigma(X_{1}) \cdot \sigma(X_{4}) \\
\vdots \\
\sigma(X_{i})^2 + \rho(X_i,X_j) \cdot \sigma(X_{i}) \cdot \sigma(X_{j}) \\
\end{bmatrix}
 $$

$$
\begin{bmatrix}
\sigma(X_{1}) \\
\sigma(X_{1}) \\
\sigma(X_{1}) \\
\vdots \\
\sigma(X_{i}) \\
\end{bmatrix}^2
+
\begin{bmatrix} 
\rho(X_1,X_2) \\
\rho(X_1,X_3) \\
\rho(X_1,X_4) \\
\vdots \\
\rho(X_i,X_j)
\end{bmatrix}

\begin{bmatrix}
\sigma(X_{1}) \\
\sigma(X_{1}) \\
\sigma(X_{1}) \\
\vdots \\
\sigma(X_{i}) \\
\end{bmatrix}

\begin{bmatrix}
\sigma(X_{2}) \\
\sigma(X_{3}) \\
\sigma(X_{4}) \\
\vdots \\
\sigma(X_{j}) \\
\end{bmatrix}
 $$

@@

>>>>

  word_document: null
documentclass: article
capsize: normalsize
geometry: margin=1in
fontsize: 11pt
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
spacing: doublespacing
bibliography: 
  - references.bib
  - knitcitations.bib
csl: cell.csl
abstract: Protein expression variation leads to phenotypic variance between cells. This has been demonstrated in cell signaling and differentiation decisions. Additionally coordinated expression of proteins between cells can tune signaling pathways either towards a more binary or analog modality. Though there is some evidence in bulk cell measurements and in bacteria that certain heteromeric subunits or metabolic pathways may be expressed in a coordinated fashion (i.e. operons in bacteria), there has not been a direct measurement of coordinated expression of proteins (independent of TFs) in metazoans (vertebrates?). Here, we measure cell-to-cell variability of relative protein abundance using quantitative proteomics of individual *Xenopus laevis* eggs and show that proteins involved in metabolic pathways or members of heteromeric complexes tend to have high correlations with other members of those pathways/complexes. Our previous work highlighted the fact that correlated expression increases the total variation of a pathway, so one would reason that certain pathways or complexes would need to compensate of this extra source of by reducing the variation of expression of these pathways or complexes. To test this we computed total variance score that took into account both the coefficient of variance and correlations between proteins in a pathway and found that the lower 10% of GO terms were highly enriched for metabolic pathways. When we looked at the relationship between CV and R between these GO terms we found a negative relationship between them, demonstrating that increased correlation needs to come at the expense of decreased variance. Simple molecular models of heteromeric complexes and metabolic pathways demonstrate that this tradeoff can result in higher efficiencies in both function and reduced energy waste. Together, our study argues for a control principle whereby coordinated expression of proteins in a pathway can require lower variance in order to reduce the overall pathway variance, enabling accurate control of active complexes or metabolic pathway activity.
---

@@

library(knitr)
library(tidyverse)
library(RColorBrewer)
#library(pheatmap)
library(ComplexHeatmap)
library(GGally)
library(VIM)
library(org.Xl.eg.db)
library(GO.db)
library(ggrepel)
library(plotly)
library(cowplot)
library(patchwork)
source("../functions.R")
library(knitcitations)
library(ggplotify)
library(latex2exp)
library(KEGG.db)
library(magick)

@@

```{r group_prots, include=FALSE, eval=FALSE}
# GO Terms
keys <- keys(org.Xl.eg.db, keytype = 'UNIPROT')

GO <- protImpute %>% dplyr::select(Uniprot) %>% dplyr::filter(Uniprot %in% keys) %>% 
  dplyr::mutate(GOterms = mapIds(org.Xl.eg.db, 
                     keys = Uniprot, 
                     keytype = 'UNIPROT', 
                     column = 'GOALL', 
                     multiVals = 'list'))

@@

allKEGG <- unnest(kegg, cols = KEGG) %>% nest(Uniprot = Uniprot) %>%
  rowwise() %>% dplyr::mutate(n = length(unlist(Uniprot))) %>% ungroup() %>%
  dplyr::mutate(name = kegg_names$kegg_name[match(KEGG,kegg_names$kegg_id)],
         type = "KEGG",
         ontology = NA) %>% rename(accession = KEGG) %>%
  dplyr::select(type, accession, name, ontology, Uniprot, n)

@@

```{r total_var, include=FALSE, eval=FALSE}
groupCor <- function(x){
  geneList <- unlist(x)

@@

allGroups <- allGroups %>% filter(!is.na(accession)) %>% rowwise() %>%
  mutate(totVar = totVarGroup(Uniprot),
         normVar = totVar / n) %>%
  filter(n > 3) %>% ungroup()

saveRDS(allGroups,"allGroups.RDS")
```

@@

```{r figure_1, echo=FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, fig.cap=fig_legend_1}
allGroups <- readRDS("allGroups.RDS")

@@

```{r figure_2, echo=FALSE, fig.height=7, fig.width=6, message=FALSE, warning=FALSE, eval=TRUE, fig.cap=fig_legend_2}
p1 <- ggplot(allGroups, aes(x = normVar, fill = interval)) + 
  geom_histogram(bins = 50) + theme_minimal() +
  #scale_fill_viridis_d()
  scale_fill_manual(values = c("#d53e4f","#3288bd","grey")) +
  xlab("Normalized Variation") + ylab("Number")

interesting_groups_low <- c("03010","00030","GO:0046496","00071",
                            "GO:0006096","GO:0015934","GO:0019843",
                            "00051","GO:0042555","04012")

interesting_groups_high <- c("GO:0042995","03008","GO:0007264","GO:0005885",
                             "GO:0007166","04150","GO:0035556","GO:0016791","GO:0007165")

interesting_groups <- c(interesting_groups_low, interesting_groups_high)

p2 <- ggplot(filter(allGroups, n > 5), aes(x = medCors, y = medCV, color = interval)) +
  geom_point(data = filter(allGroups, interval == "mid", n > 4),
             size = 2, alpha = 0.5, stroke = 0) +
  geom_point(data = filter(allGroups, interval != "mid", n > 4),
             size = 3, alpha = 0.75, stroke = 0) +
  geom_label_repel(data = filter(allGroups, accession %in% interesting_groups),
                  aes(label = name),
                  colour = 'black',
                  size = 1.5,
                  force = 5,
                  box.padding = 1) +
  #coord_cartesian(ylim=c(-10,10)) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_color_manual(values = c("#d53e4f","#3288bd","grey")) +
  xlab("Mean Correlation") + ylab("Mean Coefficient of Variation")

@@

# p5 <- rbind(mutate(allGroups, interval = "mid", name = "all"),
#       filter(allGroups, accession %in% interesting_groups)) %>%
#   dplyr::select(-Uniprot,-covs,-cv,-type,-ontology) %>%
#   unnest(cols = "cors") %>%
#   ggplot(., aes(x = fct_reorder(name,rank(interval)), y = cors, fill = interval)) +
#   #geom_boxplot(notch = FALSE, alpha = 0.75, outlier.size = 0) +
#   geom_violin() +
#   scale_fill_manual(values = c("#d53e4f","#3288bd","grey")) +
#   coord_flip() +
#   scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
#   theme_minimal() +
#   theme(text = element_text(size = 6),
#         axis.title.y=element_blank())

p6 <- p2 + plot_spacer() + plot_layout(widths = c(1,0.5))

(p1 + p3 + p4) / p6 + plot_layout(heights = c(1,5)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "none")

#p1 + p2 + p3 + plot_annotation(tag_levels = "A")

fig_legend_2 <- c("A) Histogram of the normalized variation values across all of the GO terms and KEGG pathways. The bottom and top 10% are colored blue and red respectively. B) Pairwise plot between the mean correlation and mean coefficient of variaiton within each GO term and KEGG pathway with color coding from A. Interestingly, both the top and bottom 10% of normalized variation have groups of proteins with high levels of correlation, despite high correlation adding to total variance. In the lowest 10% of normalized variation groups there appears to be a negative correlation between mean correlation and variation.")

```{r figure_3, echo=FALSE, fig.height=6, fig.width=6, message=FALSE, warning=FALSE, eval=TRUE, fig.cap=fig_legend_3}
library(circlize)

#allGroups[grep("large ribo",allGroups$name),]
prots <- allGroups %>% filter(accession == "GO:0015934") %>% pull(Uniprot) %>% unlist() %>% as.vector()
mat <- cor[prots,prots]
rownames(mat) <- mapIds(org.Xl.eg.db, keys = rownames(mat), keytype = 'UNIPROT', column = 'SYMBOL')

p1 <- as.ggplot(
  Heatmap(
    mat,
    column_title = "Large ribosomal subunit",
    column_title_gp = gpar(fontsize = 5, fontface = "bold"),
    col = colorRamp2(seq(-1,1,0.01), rev(colorRampPalette(brewer.pal(11, 'RdBu'))(201))),
    heatmap_legend_param = list(title = "R"),
    row_names_gp = gpar(fontsize = 4),
    column_names_gp = gpar(fontsize = 4),
    show_column_dend = FALSE,
    show_row_dend = FALSE,
    show_column_names = TRUE,
    show_row_names = TRUE,
    #cell_fun = function(j, i, x, y, width, height, fill) {grid.text(sprintf("%.1f", mat[i, j]), x, y, gp = gpar(fontsize = 1))},
    use_raster = FALSE
  )
)

@@

p3 <- as.ggplot(Heatmap(totVarMat,
        col = col_fun,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_column_names = FALSE,
        show_row_names = FALSE,
        border = NA,
        column_title = "R",
        row_title = "CV",
        heatmap_legend_param = list(title = "Norm Var"),
        row_names_gp = gpar(fontsize = 5),
        column_names_gp = gpar(fontsize = 5)))

bottom <- p2 + p3 + plot_layout(widths = c(1,2))
(p1 + plot_spacer()) / bottom + 
  plot_annotation(tag_levels = "A")

@@

@@

\newpage
# Supplementary material {-}

\beginsupplement
```{r sup_fig_1, echo=FALSE, fig.height=2, fig.width=6, message=FALSE, warning=FALSE, fig.cap=sup_fig_legend_1}
allGroups %>% filter(n > 4) %>%
  ggplot(., aes(x = medCors, y = medCV, color = interval)) + 
  geom_point() + 
  scale_color_manual(values = c("#d53e4f","#3288bd","grey")) +
  stat_summary(fun.data=mean_cl_normal) + 
  geom_smooth(method='lm', formula= y~x, color = "black") +
  facet_wrap(~interval, scales = "free", nrow = 1) +
  theme_minimal() +
  theme(legend.position = "none")
